---
title: "Constraining GLM trends from MMEs using observations: 3 regression parameter example"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(cmdstanr)
library(bayesplot)
library(glmmTMB)
```

# Introduction

Let's repeat the work from example.rmd but instead using a length 3 vector for $b^j$, $j = 1,...,m$ and $a$.

Instead of just considering a Poisson trend with an intercept and slope, lets consider a Gamma distribution with an intercept, slope and shape.

Then the observations are distributed as a Gamma glm:

\begin{equation}
    z_t | \nu(t) \sim \text{Gamma}(\alpha_z, \beta_z(t)), \qquad t = t_1, ... t_{n_h},
\end{equation}
\begin{equation}
    log(\frac{\alpha_z}{ \beta_z(t)}) = a_1 + a_2 t,
\end{equation}
\begin{equation}
    log(\alpha_z) = a_3.
\end{equation}

And similarly for the climate simulations:

\begin{equation}
    x^j_t | \nu(t) \sim \text{Gamma}(\alpha_{x^j}, \beta_{x^j}(t)), \qquad t = t_1, ... t_{n_h},
\end{equation}
\begin{equation}
    log(\frac{\alpha_{x^j}}{ \beta_{x^j}(t)}) = b^j_1 + b^j_2 t,
\end{equation}
\begin{equation}
    log(\alpha_{x^j}) = b^j_3.
\end{equation}

# The similarity parameter in N-dimensions

As the vector $\delta$ gets longer, the framework becomes progressively more difficult to fit in stan. 
In our testing for len(delta) > 2, we get divergent transitions and maximum treedepth issues.
It is also harder to visualise and interpret higher dimensional vectors.

As an alternative to having $\delta$ as a vector we could assume that all components of delta are the same and replace it with a single scalar determining the similarity between the ensemble and the real world.
Then:
\begin{equation}
    \Gamma = \frac{1-\delta}{\delta} \text{diag}(\Sigma).
\end{equation}

This corresponds to the belief that all of the components of $a$ have the same ratio of variances with the ensemble and removes the ability to consider one component as exchangeable with the ensemble but another as independent, for example.

# Simulated data

Again set up the time covariate, t in years, which we rescale to t_sc and generate some simulated parameters (intercept and slope and log(shape)) where we know the true $\Sigma$, $\mu$ and $w$.
In this example we assume we have a longer observational period.

```{r}
start_date <- 1850
end_hist <- 2014
end_date <- 2100
nh <- end_hist - start_date + 1
nf <- end_date - start_date + 1
t <- seq(start_date, end_date, length.out = nf)
t_sc <- (t - start_date)/(nf - 1)
```

```{r}
m = 20 # number of simulations
set.seed(1234)

Sigma_true <- cbind(c(.02,-.01, 0),c(-.01,.2, 0), c(0, 0, 0.1))
mu_true <- cbind(2, 0, 0)
L_true <- t(chol(Sigma_true))

 # The true simulation regression parameters
b_true <- matrix(rep(mu_true, m), nrow = length(mu_true)) + L_true %*% matrix(rnorm(m * length(mu_true)), nrow = length(mu_true))
w_true <- c(0.5,-0.3, 0)

# The true real world regression parameters
a_true <- as.numeric(mu_true) + L_true %*% rnorm(length(mu_true))# + w_true 

print("mu is:")
print(mu_true)
print("Sigma is:")
print(Sigma_true)
```

Now lets simulate the data as a Gamma distributed GLM.

```{r}
B <- cbind(rep(1, nf), t_sc)

X <- matrix(rgamma(n = m * nf, exp(b_true[3,]), exp(b_true[3,])/exp(as.numeric(B %*% b_true[1:2,]))), ncol = m)

z <- rgamma(nh, exp(a_true[3]), exp(a_true[3])/exp(as.numeric(B[1:nh,] %*% a_true[1:2])))

data_df <- data.frame(X) %>%
    mutate(t_sc = t_sc, t = t) %>%
    pivot_longer(-c(t_sc, t),names_to="model", names_prefix="X") %>%
    bind_rows(
        data.frame(value = z, t_sc = t_sc[1:nh], t = t[1:nh], model = "obs")
    )
```

```{r}
data_df %>%
    ggplot(aes(x = t, y = value)) +
    geom_line() +
    facet_wrap(~model)
```

# Fitting the model

As in the previous example we do a maximum likelihood step.
However as we need the shape parameter to be fitted using a log link we use the glmmTMB package instead of the base R glm function.
For the Gamma distribution using the glmmTMB a log link for the shape (dispersion) parameter is the default.

```{r}
fit_list <- list()
for(j in 1:m){
        fit_list[[j]] <- glmmTMB(
            value ~ t_sc,
            family=Gamma(link = "log"),
            filter(data_df, model == j)
        )
}
a_fit <- glmmTMB(value ~ t_sc, family=Gamma(link = "log"), filter(data_df, model == "obs"))
```

```{r}
b_hat <- matrix(ncol = length(mu_true), nrow = m)
J <- array(0, dim = c(m, length(mu_true), length(mu_true)))
for(j in 1:m){
    bfit <- fit_list[[j]]
    b_hat[j,] <- c(unname(fixef(bfit)$cond),unname(fixef(bfit)$disp))
    J[j,,] <- t(chol(vcov(fit_list[[j]], full = TRUE)))
}
a_hat <- c(unname(fixef(a_fit)$cond),unname(fixef(a_fit)$disp))
J_a <- t(chol(vcov(a_fit, full = TRUE)))
```

# Different values of delta

Consider a number of different $\delta$ values.
Note the model cannot accept a value of $\delta = 0$.

```{r}
deltas = c(0.001,seq(.25,1,.25))
```

```{r}
prior_hyper_param = numeric()              # an empty vector will result in no priors
# prior_hyper_param = c(
#     mu_mu_1 = 0,
#     mu_mu_2 = 0,
#     mu_mu_3 = 0,
#     mu_sd_1 = 10,
#     mu_sd_2 = 5,
#     mu_sd_3 = 5,
#     eta = 2,
#     gam = 2.5
# )    # a vector of prior hyper parameters

data_list_mult_delta <- list(
    len_a = 3,
    m = m,
    b_hat = b_hat,
    a_hat = a_hat,
    J = J,
    J_a = J_a,
    len_hyper_param = length(prior_hyper_param),
    hyper_param = prior_hyper_param,
    n_delta = length(deltas),
    delta = deltas                                 # Fixed values of delta
)
```


```{r}
# This stan model fits the hierarchical model for a set of fixed deltas.
mod <- cmdstan_model("stan/glm_constr_mme_obs_fixed_1D_delta_vector.stan")

data_list_mult_delta[["delta"]] <- deltas
data_list_mult_delta[["n_delta"]] <- length(deltas)

fit_fixed_delta_mult_delta <- mod$sample(
    data = data_list_mult_delta,
    seed = 1234,
    iter_warmup = 1000,
    iter_sampling = 1000,
    chains = 4,
    parallel_chains = 4,
    refresh = 0,
    adapt_delta=.95,
    show_exceptions = FALSE
)
```

Lets use precentage change from the start of the time period to the end of the time period (t_sc = 0, t = 1850 to t_sc = 1, t = 2100) as an interesting example: 

\begin{align}
    100\%\frac{E(z|t = 2100) - E(z|t = 1850)}{E(z|t = 1850)} &=  100\%\frac{\text{exp}(a_1 + a_2) - \text{exp}(a_1)}{\text{exp}(a_1)} \\
    &= 100\%(\text{exp}(a_2) - 1) 
\end{align}

Plotting for the different choices of $\delta$

```{r, echo=FALSE}
fit_fixed_delta_mult_delta$draws(c("a"), format="df") %>% 
    pivot_longer(contains("["), values_to = "a", names_to = c("index","comp"), names_pattern = "\\[(.+)\\,(.+)\\]", names_transform=list(index = as.integer,comp = as.integer)) %>%
    mutate(delta = deltas[index]) %>%
    filter(comp == 2) %>%
    mutate(pcnt_diff = 100 * (exp(a) - 1)) %>%
    ggplot() +
    geom_density(aes(x = pcnt_diff)) +
    # xlim(-100,100) +
    labs(x = "% difference between start and end", y = NULL) +
    facet_wrap(~delta)

```

Again this can be done with a prior on $\delta$, e.g. $\delta \sim \text{Unif}(0,1)$.

```{r}
prior_hyper_param = numeric()              # an empty vector will result in no priors
# prior_hyper_param = c(
#     mu_mu_1 = 0,
#     mu_mu_2 = 0,
#     mu_sd_1 = 10,
#     mu_sd_2 = 5,
#     eta = 2,
#     gam = 2.5
# )    # an length 6 vector of prior hyper parameters

data_list <- list(
    len_a = 3,
    m = m,
    b_hat = b_hat,
    a_hat = a_hat,
    J = J,
    J_a = J_a,
    len_hyper_param = length(prior_hyper_param),
    hyper_param = prior_hyper_param,
    delta_hyper_param = c(1,1)          # Prior parameters for delta ~ Beta() prior
)
```

```{r}
mod <- cmdstan_model("stan/glm_constr_mme_obs_1D_delta.stan")

fit_prior_on_delta <- mod$sample(
    data = data_list,
    seed = 1234,
    iter_warmup = 1000,
    iter_sampling = 1000,
    chains = 4,
    parallel_chains = 4,
    refresh = 0,
    show_exceptions = FALSE
)
```

```{r, echo=FALSE}
fit_prior_on_delta$draws(c("a", "b_mme", "a_ind"), format = "df") %>%
    pivot_longer(contains("["), names_to = c("param", "comp"), names_pattern = "(.+)\\[(.)\\]") %>%
    mutate(comp = c("Intercept","Slope","Intercept (shape)")[as.integer(comp)]) %>%
    ggplot(aes(x = value)) +
    geom_density(aes(colour = param, fill = param), alpha = .2) +
    labs(x = NULL, y = NULL) +
    scale_colour_manual(name = NULL, values = c(`a` = "blue", `a_ind` = "orange", `b_mme` = "black")) +
    scale_fill_manual(name = NULL, values = c(`a` = "blue", `a_ind` = "orange", `b_mme` = "black")) +
    facet_wrap(~comp, scales = "free_x")
```

```{r, echo=FALSE}
mcmc_areas(fit_prior_on_delta$draws("delta"))
```

